### 2013-08-23 doc/dissertation/document/chapters/mc_background/mc_background.tex
spell check for spn chapter edits
moving subomain nuemann ulam results for multiple sets into an appendix

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Method Selection}
\label{sec:mc_method_selection}
The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
to estimate the error in a residual Monte Carlo solve instead of the
direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
effectiveness of the adjoint method over the direct method within the
context of MCSA, we choose the two-dimensional time-dependent Poisson
equation as a simple model transport problem\footnote{The Poisson
  equation is a simple form of the diffusion equation and has a
  similar elliptic character in its time-dependent form to the neutron
  transport equations that will be solved in the next chapter.}:
\begin{equation}
  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five-point stencil,
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine-point stencil,
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method with the collision estimator as
compared to the direct method. To assess both the CPU time and number
of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of the number of mesh
elements, fixing the spectral radius of the system at a constant value
for each variation. Both the five-point and nine-point stencils were
used with both the direct and adjoint solvers. For each case, $N
\times N$ total random walk permutations were computed per MCSA
iteration where $N \times N$ is the number of discrete grid points (or
degrees of freedom) in the system. Solver parameters were set to a
weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds, and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance, both as a function of the problem size. All
computations presented in this section and the remaining sections of
this chapter were completed on a 3.0 GHz Intel Core 2 Quad Q9650 CPU
machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in a speedup over the direct solver
while the number of iterations required to converge is also reduced as
shown in Figure~\ref{fig:poisson_iterations}. We expect this for
several reasons. First, with an equivalent number of histories
specified for both solvers per MCSA iteration and a system of size $N
\times N$, the direct solver will compute a single random walk for
each state in the system per iteration to acquire a solution in that
state. This is necessary in the direct method to ensure a contribution
from each state as the random walk sequence will only contribute to
the starting state. For the adjoint method, a total of $N \times N$
random walk events will have their starting state determined by
sampling the residual vector. Because the random walk sequence
contributes to the state in which it currently resides, sampling the
residual vector as the Monte Carlo source gives a higher density of
random walk events in regions with a high residual, thus giving a more
accurate correction in that region due to reduced statistical
error. From an iteration perspective,
Figure~\ref{fig:poisson_iterations} shows that using the direct method
yields a roughly unchanging number of iterations required to converge
as the problem size increases. Again, if we desire a correction value
for all states in the problem, then we must start a random walk in
each state in the system without taking into account the structure of
the residual vector. Because of this, as the problem size grows adding
histories is ineffective as many are added in states where the error
is smaller than in other parts of the problem, ultimately not reducing
the number of iterations needed to converge. Conversely, as the
problem size grows in the adjoint method, the additional stochastic
histories that will be computed are concentrated in regions with a
large residual, further reducing the stochastic error in the
correction in those regions and subsequently reducing the required
number of iterations to converge.

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
gives the results of these computations. First, it is worthy to note
on the semi-log plot that we are indeed achieving the expected
exponential convergence from MCSA with both Monte Carlo
solvers. Second, we note that using the adjoint method with the same
number of stochastic histories per MCSA iteration gives a faster rate
of converge for the same reasons as above. We also note here that
fewer iterations are required for convergence when the 9-point stencil
is used to discretize the Laplacian operator (although at no gain in
speed as given by the results in
Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
smaller discretization error directly corresponds to a more well
defined residual source generated by the Richardson extrapolation for
the Monte Carlo calculation. In addition, the better defined source is
transported through a domain described more accurately by the 9-point
stencil, thus yielding a more accurate correction vector from the
Monte Carlo calculation.
\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the adjoint and direct solvers are used with the five
      point and nine point stencils. A higher rate of convergence is
      observed for MCSA using the adjoint Monte Carlo solver as
      compared to the direct method when both solvers compute the same
      number of random walks per iteration.}}
  \label{fig:poisson_convergence}
\end{figure}

%%---------------------------------------------------------------------------%%
\section{MCSA Comparison to Sequential Monte Carlo}
\label{subsec:sequential_comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we
compare its performance to Halton's Sequential Monte Carlo method on
which previous work in this area was based. For this comparison, we
use the same transient Poisson problem as described in the previous
section and choose only the 5-point stencil to discretize the
Laplacian operator as the previous results yielded little qualitative
difference between the discretizations. Both MCSA and Halton's method
are used with the adjoint Monte Carlo solver and the collision
estimator. In order to complete the same study as in the previous
section, the number of histories computed by the Monte Carlo solver at
each iteration had to be doubled to $2 \times N \times N$ in order to
ensure convergence in Sequential Monte Carlo Method. For the majority
of the problems in the previous section, the Sequential method used
with $N \times N$ histories would not
converge.

@@

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Parameter and Estimator Analysis}
\label{sec:parameter_estimator_analysis}
With the adjoint method shown to be more effective than the direct
method and MCSA to have better iterative and timing performance than
sequential Monte Carlo, we now aim to study the effects of the adjoint
Monte Carlo parameters, weight cutoff and number of histories, on MCSA
performance with both the collision and expected value
estimators. With the same Poisson problem, we will use a $200 \times
200$ grid and a convergence tolerance of \sn{1}{-8} for all
calculations. To study the effects of the number of Monte Carlo
histories per MCSA iteration, the weight cutoff was fixed at
\sn{1}{-4} while the number of histories per iteration was varied from
6,000 to 100,000. It was observed that MCSA would not converge for
this problem using the collision estimator with less than 6,000
histories while the expected value estimator permitted convergence
with only 2,000 histories.

@@

In this chapter, the fundamentals of Monte Carlo Synthetic
Acceleration methods were presented and explored in the context of a
simple model transport system. The following are the significant
observations and findings.

>>>>

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration Analysis}
\label{sec:mcsa_analysis}
In this section we analyze the effects of various MCSA parameters to
better our understanding of the algorithm and facilitate future
work. In addition, we demonstrate the effectiveness of MCSA as
compared to Halton's Sequential Monte Carlo method. To do this, we
choose the two-dimensional time-dependent Poisson equation as a simple
model transport problem\footnote{The Poisson equation is a simple form
  of the diffusion equation and has a similar elliptic character in
  its time-dependent form to the neutron transport equations that will
  be solved in the next chapter.}:
\begin{equation}
  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five-point stencil,
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine-point stencil,
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

\subsection{Monte Carlo Method Selection}
\label{sec:mc_method_selection}
The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
to estimate the error in a residual Monte Carlo solve instead of the
direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
effectiveness of the adjoint method over the direct method within the
context of MCSA, we solve Poisson equation in a series of numerical
experiments. A timing and convergence study is used to demonstrate
the effectiveness of the adjoint method with the collision estimator
as compared to the direct method. To assess both the CPU time and
number of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of the number of mesh
elements, fixing the spectral radius of the system at a constant value
for each variation. Both the five-point and nine-point stencils were
used with both the direct and adjoint solvers. For each case, $N
\times N$ total random walk permutations were computed per MCSA
iteration where $N \times N$ is the number of discrete grid points (or
degrees of freedom) in the system. Solver parameters were set to a
weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds, and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance, both as a function of the problem size. All
computations presented in this section and the remaining sections of
this chapter were completed on a 3.0 GHz Intel Core 2 Quad Q9650 CPU
machine with 16 GB 1067 MHz DDR3 memory.

\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in a speedup over the direct solver
while the number of iterations required to converge is also reduced as
shown in Figure~\ref{fig:poisson_iterations}. We expect this for
several reasons. First, with an equivalent number of histories
specified for both solvers per MCSA iteration and a system of size $N
\times N$, the direct solver will compute a single random walk for
each state in the system per iteration to acquire a solution in that
state. This is necessary in the direct method to ensure a contribution
from each state as the random walk sequence will only contribute to
the starting state. For the adjoint method, a total of $N \times N$
random walk events will have their starting state determined by
sampling the residual vector. Because the random walk sequence
contributes to the state in which it currently resides, sampling the
residual vector as the Monte Carlo source gives a higher density of
random walk events in regions with a high residual, thus giving a more
accurate correction in that region due to reduced statistical
error.

From an iteration perspective, Figure~\ref{fig:poisson_iterations}
shows that using the direct method yields a roughly unchanging number
of iterations required to converge as the problem size
increases. Again, if we desire a correction value for all states in
the problem, then we must start a random walk in each state in the
system without taking into account the structure of the residual
vector. Because of this, as the problem size grows adding histories is
ineffective as many are added in states where the error is smaller
than in other parts of the problem, ultimately not reducing the number
of iterations needed to converge. Conversely, as the problem size
grows in the adjoint method, the additional stochastic histories that
will be computed are concentrated in regions with a large residual,
further reducing the stochastic error in the correction in those
regions and subsequently reducing the required number of iterations to
converge.

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
gives the results of these computations. We note that using the
adjoint method with the same number of stochastic histories per MCSA
iteration gives a faster rate of converge for the same reasons as
above. We also note here that fewer iterations are required for
convergence when the 9-point stencil is used to discretize the
Laplacian operator (although at no gain in speed as given by the
results in Figure~\ref{fig:poisson_cpu_time}). This is due to the fact
that the smaller discretization error directly corresponds to a more
well defined residual source generated by the Richardson extrapolation
for the Monte Carlo calculation. In addition, the better defined
source is transported through a domain described more accurately by
the 9-point stencil, thus yielding a more accurate correction vector
from the Monte Carlo calculation.

\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the adjoint and direct solvers are used with the five
      point and nine point stencils. A higher rate of convergence is
      observed for MCSA using the adjoint Monte Carlo solver as
      compared to the direct method when both solvers compute the same
      number of random walks per iteration.}}
  \label{fig:poisson_convergence}
\end{figure}

\subsection{MCSA Comparison to Sequential Monte Carlo}
\label{subsec:sequential_comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we
compare its performance to Halton's Sequential Monte Carlo. For this
comparison, we use the same transient Poisson problem as described in
the previous section and choose only the 5-point stencil to discretize
the Laplacian operator as the previous results yielded little
qualitative difference between the discretizations. Both MCSA and
Halton's method are used with the adjoint Monte Carlo solver and the
collision estimator. In order to complete the same study as in the
previous section, the number of histories computed by the Monte Carlo
solver at each iteration had to be doubled to $2 \times N \times N$ in
order to ensure convergence in Sequential Monte Carlo Method. For the
majority of the problems in the previous section, the Sequential
method used with $N \times N$ histories would not converge.

@@

\subsection{Monte Carlo Parameter and Estimator Analysis}
\label{sec:parameter_estimator_analysis}
We now study the effects of the adjoint Monte Carlo parameters, weight
cutoff and number of histories, on MCSA performance with both the
collision and expected value estimators. With the same Poisson
problem, we will use a $200 \times 200$ grid and a convergence
tolerance of \sn{1}{-8} for all calculations. To study the effects of
the number of Monte Carlo histories per MCSA iteration, the weight
cutoff was fixed at \sn{1}{-4} while the number of histories per
iteration was varied from 6,000 to 100,000. It was observed that MCSA
would not converge for this problem using the collision estimator with
less than 6,000 histories while the expected value estimator permitted
convergence with only 2,000 histories.

@@

In this chapter, the fundamentals of Monte Carlo Synthetic
Acceleration methods were presented and explored in the context of a
simple model transport system. The following are the significant
observations, findings, and contributions.
